dyngen is a multi-modal simulator for spearheading future single-cell omics analyses

### 2021-05-20 manuscript/v5/manuscript.tex
update

\usepackage{hyperref}
\hypersetup{
  pdftitle={dyngen: a multi-modal simulator for spearheading new single-cell omics analyses},
  pdfauthor={Robrecht Cannoodt; Wouter Saelens; Louise Deconinck; Yvan Saeys},
  hidelinks}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}

\title{dyngen: a multi-modal simulator for spearheading new single-cell omics analyses}
\author{Robrecht Cannoodt\textsuperscript{$\dagger{}$,1,2,3} \and Wouter Saelens\textsuperscript{$\dagger{}$,1,2,4} \and Louise Deconinck\textsuperscript{1,2} \and Yvan Saeys\textsuperscript{1,2,*}}
\date{12 May 2021}

@@

\section{Abstract}\label{abstract}

We present dyngen, a novel, multi-modal simulation engine for studying
dynamic cellular processes at single-cell resolution. dyngen is more
flexible than current single-cell simulation engines, and allows better
method development and benchmarking, thereby stimulating development and
testing of novel computational methods. We demonstrate its potential for
spearheading novel computational methods on three novel applications:
aligning cell developmental trajectories, cell-specific regulatory
network inference and estimation of RNA velocity.

\section{Introduction}\label{introduction}

@@

Another, more experimental use of simulation engines is their important
role in spearheading the development of novel computational methods,
possibly even before real data is available. In this way, simulation
engines can be used to assess the value of novel experimental protocols
or treatments. Simulation engines are also increasingly important when
it comes to finding alternatives to animal models, for example for drug
testing and precision medicine. In such scenarios, cellular simulations
can act as digital twins, offering unlimited experimentation \emph{in
silico} \cite{bjornsson_digitaltwinspersonalize_2019}.

Simulating realistic data requires that the underlying biology is
recapitulated as best as possible, and in the case of transcriptomics
data this typically involves modelling the underlying gene regulatory
networks. Simulators of ``bulk'' microarray or RNA-sequencing profiles
simulate biological processes (e.g.~transcription, translation) by
translating a database of known regulatory interactions into a set of
ordinary differential equations (ODE)
\cite{roy_systemgeneratingtranscription_2008,hache_gengesystematicgeneration_2009,schaffter_genenetweaversilicobenchmark_2011,vandenbulcke_syntrengeneratorsynthetic_2006}.
These methods have been instrumental in performing benchmarking studies
\cite{prill_rigorousassessmentsystems_2010,marbach_revealingstrengthsweaknesses_2010,marbach_wisdomcrowdsrobust_2012}.
However, the advent of single-cell omics introduced several new types of
analyses (e.g.~trajectory inference, RNA velocity, cell-specific network
inference) which exploit the higher resolution of single-cell versus
bulk omics \cite{luecken_currentbestpractices_2019}. In addition,
the data characteristics of single-cell omics are vastly different from
bulk omics, typically having much lower library sizes and a higher
dropout rate, but also a high number of profiles
\cite{vallejos_normalizingsinglecellrna_2017}. The low library
sizes, in particular, are problematic as ODEs are ill-suited for
performing low-molecule simulations
\cite{gillespie_exactstochasticsimulation_1977}. This necessitates
the development of new single-cell simulators.

@@

\begin{figure}[htbp]
    \centering
    \includegraphics[width=\linewidth]{figure_1.pdf}
    \caption{
      \textbf{Showcase of dyngen functionality.}
      \textbf{A:} Changes in abundance levels are driven strictly by gene regulatory reactions.
      \textbf{B:} The input Gene Regulatory Network (GRN) is defined such that it models a dynamic process of interest.
      \textbf{C:} The reactions define how abundance levels of molecules change at any particular time point.
      \textbf{D:} Firing many reactions can significantly alter the cellular state over time.
      \textbf{E:} dyngen keeps track of the likelihood of a reaction firing during small intervals of time, called the propensity, as well as the actual number of firings.
      \textbf{F:} Similarly, dyngen can also keep track of the regulatory activity of every interaction.
      \textbf{G:} A benchmark of trajectory inference methods has already been performed using the cell state ground-truth \cite{saelens_comparisonsinglecelltrajectory_2019}.
      \textbf{H:} The cell state ground-truth enables evaluating trajectory alignment methods.
      \textbf{I:} The reaction propensity ground-truth enables evaluating RNA velocity methods.
      \textbf{J:} The cellwise regulatory network ground-truth enables evaluating cell-specific gene regulatory network inference methods.
    }
    \label{fig:overview}
\end{figure}

\section{Results}\label{results}

@@

\begin{figure}[htbp]
    \centering
    \includegraphics[width=\linewidth]{figure_2.pdf}
    \caption{
      \textbf{dyngen provides ground-truth data for a variety of applications (left), which can be used to quantitatively evaluate methods (right).} Box plots denote the $Q_0$ to $Q_4$ quartile values.
      \textbf{A:} Trajectory alignment aligns two trajectories between samples. We evaluate dynamic time warping (DTW) and cellAlign when aligning two linear trajectories with different kinetic parameters based on the area differences between the worst possible alignment and the predicted alignment (Area Between Worst And Prediction, or ABWAP). 
      \textbf{B:} RNA velocity calculates for each cell the direction in which the expression of each gene is moving. We evaluated scVelo and velocyto by comparing these vectors with the known velocity vector (velocity correlation) and with the known direction of the cellular trajectory in a dimensionality reduction (velocity arrow cosine).
      \textbf{C:} Cell-specific network inference (CSNI) predicts the regulatory network of every individual cell. We evaluate each cell-specific regulatory network with typical metrics for network inference: the Area Under the Receiver Operating Characteristics-curve (AUROC) and Area Under the Precision Recall-curve (AUPR). We evaluate three CSNI methods by computing the mean AUROC and AUPR across all cells.
    }
    \label{fig:applications}
\end{figure}

@@

\section{Discussion}\label{discussion}

dyngen's single-cell simulations can be used to evaluate common
single-cell omics computational methods such as clustering, batch
correction, trajectory inference, and network inference. However, the
framework is flexible enough to be adaptable to a broad range of
applications, including methods that integrate clustering, network
inference, and trajectory inference. In this respect, dyngen may promote
the development of new tools in the single-cell field similarly as other
simulators have done in the past
\cite{schaffter_genenetweaversilicobenchmark_2011,ewing_combiningtumorgenome_2015}.
Additionally, one could anticipate technological developments in
single-cell multi-omics. In this way, dyngen allows designing and
evaluating the performance and robustness of new types of computational
analyses before experimental data becomes available, comparing which
experimental protocol is the most cost-effective in producing
qualitative and robust results in downstream analysis. One major
assumption of dyngen is that cells are assumed to be well-mixed and
independent from each other. Subdividing a cell into multiple 2D or 3D
subvolumes or allowing cells to exchange molecules, respectively, could
pave the way to better study key cellular processes such as cell
division, intercellular communication, and migration
\cite{smith_spatialstochasticintracellular_2019}.

@@

\section{Methods}\label{sec:dyngen-methods}

@@

\subsection{Defining the module network}\label{sec:dyngen-modules}

@@

\subsection{Generating the gene regulatory network}\label{sec:dyngen-grn}

@@

\subsection{Convert gene regulatory network to a set of reactions}\label{sec:dyngen-reactions}

@@

\subsubsection{\texorpdfstring{Propensity of transcription when \(N=1\)}{Propensity of transcription when N=1}}\label{propensity-of-transcription-when-n1}

@@

\subsubsection{\texorpdfstring{Propensity of transcription when \(N=2\)}{Propensity of transcription when N=2}}\label{propensity-of-transcription-when-n2}

@@

\subsubsection{\texorpdfstring{Propensity of transcription for an arbitrary \(N\)}{Propensity of transcription for an arbitrary N}}\label{propensity-of-transcription-for-an-arbitrary-n}

@@

\subsubsection{\texorpdfstring{Propensity of transcription for a large \(N\)}{Propensity of transcription for a large N}}\label{propensity-of-transcription-for-a-large-n}

@@

\subsubsection{Independence, synergism and basal expression}\label{independence-synergism-and-basal-expression}

@@

\subsection{Simulate single cells}\label{dyngen-simcell}

@@

\subsection{Simulate experiment}\label{sec:dyngen-experiment}

From the SSA simulation we obtain the abundance levels of all the
molecules at every state. We need to replicate technical effects
introduced by experimental protocols in order to obtain data that is
similar to real data. For this, the cells are sampled from the
simulations and molecules are sampled for each of the cells. Gene
capture rates and library sizes are empirically derived from real
datasets as to match real technical variation.

\subsubsection{Sample cells}\label{sample-cells}

@@

\subsubsection{Sample molecules}\label{sample-molecules}

@@

\subsection{Simulating batch effects}\label{sec:dyngen-batcheffect}

@@

\subsection{Determining the ground-truth trajectory}\label{sec:dyngen-groundtruth}

@@

\subsection{Determining the cell-specific ground-truth regulatory network}\label{sec:dyngen-extractgrn}

Calculating the regulatory effect of a regulator \(R\) on a target \(T\)
(Supplementary Fig.~4F) requires determining the contribution of \(R\)
in the propensity function of the transcription of \(T\)
(section~\ref{sec:dyngen-reactions}) with respect to other regulators.
This information is useful, amongst others, for benchmarking
cell-specific network inference methods.

@@

\subsection{Comparison of cell-specific network inference methods}\label{sec:dyngen-nicompare}

42 datasets were generated using the 14 different predefined backbones
and three different seeds. For every cell in the dataset, the
transcriptomics profile and the corresponding cell-specific ground-truth
regulatory network was determined (Section~\ref{sec:dyngen-extractgrn}).

@@

\subsection{Comparison of RNA velocity methods}\label{sec:dyngen-velcompare}

@@

 \subsection{Comparison of trajectory alignment}\label{sec:dyngen-tacompare}

@@

\section{Data Availability}\label{data-availability}

@@

\section{Code Availability}\label{code-availability}

@@

\section{Author contributions}\label{author-contributions}

@@

\begin{thebibliography}{10}

	\bibitem{aibar_scenicsinglecellregulatory_2017}
	Sara Aibar, Carmen~Bravo Gonz{\a'a}lez-Blas, Thomas Moerman, V{\^a}n~Anh
	Huynh-Thu, Hana Imrichova, Gert Hulselmans, Florian Rambow, Jean-Christophe
	Marine, Pierre Geurts, Jan Aerts, Joost {van den Oord}, Zeynep~Kalender Atak,
	Jasper Wouters, and Stein Aerts.
	\newblock {{SCENIC}}: Single-cell regulatory network inference and clustering.
	\newblock {\em Nature Methods}, 10 2017.

	\bibitem{alpert_alignmentsinglecelltrajectories_2018}
	Ayelet Alpert, Lindsay~S. Moore, Tania Dubovik, and Shai~S. Shen-Orr.
	\newblock Alignment of single-cell trajectories to compare cellular expression
	dynamics.
	\newblock {\em Nature Methods}, 15(4):267--270, 4 2018.

	\bibitem{bergen_generalizingrnavelocity_2020}
	Volker Bergen, Marius Lange, Stefan Peidli, F.~Alexander Wolf, and Fabian~J.
	Theis.
	\newblock Generalizing {{RNA}} velocity to transient cell states through
	dynamical modeling.
	\newblock {\em Nature Biotechnology}, 38(12):1408--1414, 12 2020.

	\bibitem{bjornsson_digitaltwinspersonalize_2019}
	Bergthor Bj{\"o}rnsson, Carl Borrebaeck, Nils Elander, Thomas Gasslander,
	Danuta~R. Gawel, Mika Gustafsson, Rebecka J{\"o}rnsten, Eun~Jung Lee, Xinxiu
	Li, Sandra Lilja, David Mart\'inez-Enguita, Andreas Matussek, Per
	Sandstr{\"o}m, Samuel Sch{\"a}fer, Margaretha Stenmarker, X.~F. Sun, Oleg
	Sysoev, Huan Zhang, Mikael Benson, and {on behalf of the Swedish Digital Twin
		Consortium}.
	\newblock Digital twins to personalize medicine.
	\newblock {\em Genome Medicine}, 12(1):4, 12 2019.

	\bibitem{cacchiarelli_aligningsinglecelldevelopmental_2018}
	Davide Cacchiarelli, Xiaojie Qiu, Sanjay Srivatsan, Anna Manfredi, Michael
	Ziller, Eliah Overbey, Antonio Grimaldi, Jonna Grimsby, Prapti Pokharel,
	Kenneth~J. Livak, Shuqiang Li, Alexander Meissner, Tarjei~S. Mikkelsen,
	John~L. Rinn, and Cole Trapnell.
	\newblock Aligning {{Single}}-{{Cell Developmental}} and {{Reprogramming
			Trajectories Identifies Molecular Determinants}} of {{Myogenic Reprogramming
			Outcome}}.
	\newblock {\em Cell Systems}, 7(3):258--268.e3, 9 2018.

	\bibitem{cannoodt_singlecellomicsdatasets_2018}
	Robrecht Cannoodt, Wouter Saelens, Helena Todorov, and Yvan Saeys.
	\newblock Single-cell -omics datasets containing a trajectory.
	\newblock {\em Zenodo}, 10 2018.

	\bibitem{conover_multiplecomparisonsprocedures_1979}
	William~Jay Conover and Ronald~L Iman.
	\newblock On multiple-comparisons procedures.
	\newblock Technical report, {Technical report, Los Alamos Scientific
		Laboratory}, 1979.

	\bibitem{duo_systematicperformanceevaluation_2018}
	Angelo Du{\a`o}, Mark~D. Robinson, and Charlotte Soneson.
	\newblock A systematic performance evaluation of clustering methods for
	single-cell {{RNA}}-seq data.
	\newblock {\em F1000Research}, 7:1141, 2018.

	\bibitem{ewing_combiningtumorgenome_2015}
	Adam~D. Ewing, Kathleen~E. Houlahan, Yin Hu, Kyle Ellrott, Cristian Caloian,
	Takafumi~N. Yamaguchi, J.~Christopher Bare, Christine P'ng, Daryl Waggott,
	Veronica~Y. Sabelnykova, {ICGC-TCGA DREAM Somatic Mutation Calling Challenge
		Participants}, Liu Xi, Ninad Dewal, Yu~Fan, Wenyi Wang, David Wheeler,
	Andreas Wilm, Grace~Hui Ting, Chenhao Li, Denis Bertrand, Niranjan Nagarajan,
	Qing-Rong Chen, Chih-Hao Hsu, Ying Hu, Chunhua Yan, Warren Kibbe, Daoud
	Meerzaman, Kristian Cibulskis, Mara Rosenberg, Louis Bergelson, Adam Kiezun,
	Amie Radenbaugh, Anne-Sophie Sertier, Anthony Ferrari, Laurie Tonton, Kunal
	Bhutani, Nancy~F. Hansen, Difei Wang, Lei Song, Zhongwu Lai, Yang Liao, Wei
	Shi, Jos{\a'e} Carbonell-Caballero, Joaqu\'in Dopazo, Cheryl C.~K. Lau,
	Justin Guinney, Michael~R. Kellen, Thea~C. Norman, David Haussler, Stephen~H.
	Friend, Gustavo Stolovitzky, Adam~A. Margolin, Joshua~M. Stuart, and Paul~C.
	Boutros.
	\newblock Combining tumor genome simulation with crowdsourcing to benchmark
	somatic single-nucleotide-variant detection.
	\newblock {\em Nature Methods}, 12(7):623--630, 7 2015.

	\bibitem{ferrell_bistabilitybifurcationswaddington_2012}
	James~E Ferrell.
	\newblock Bistability, {{Bifurcations}}, and {{Waddington}}'s {{Epigenetic
			Landscape}}.
	\newblock {\em Current Biology}, 22(11):R458--R466, 6 2012.

	\bibitem{gibson_probabilisticmodelprokaryotic_2000}
	Michael~A. Gibson and Jehoshua Bruck.
	\newblock A probabilistic model of a prokaryotic gene and its regulation.
	\newblock {\em Computational Methods in Molecular Biology: From Genotype to
		Phenotype, MIT press, Cambridge}, 2000.

	\bibitem{gillespie_exactstochasticsimulation_1977}
	Daniel~T. Gillespie.
	\newblock Exact stochastic simulation of coupled chemical reactions.
	\newblock {\em The Journal of Physical Chemistry}, 81(25):2340--2361, 12 1977.

	\bibitem{giorgino_computingvisualizingdynamic_2009}
	Toni Giorgino.
	\newblock Computing and {{Visualizing Dynamic Time Warping Alignments}} in
	{{R}}: {{The}} dtw {{Package}}.
	\newblock {\em Journal of Statistical Software}, (7), 9 2009.

>>>>

\usepackage{hyperref}
\hypersetup{
  pdftitle={dyngen is a multi-modal simulator for spearheading future single-cell omics analyses},
  pdfauthor={Robrecht Cannoodt; Wouter Saelens; Louise Deconinck; Yvan Saeys},
  hidelinks}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage[numbers]{natbib}

\title{dyngen is a multi-modal simulator for spearheading future single-cell omics analyses}
\author{Robrecht Cannoodt\textsuperscript{$\dagger{}$,1,2,3} \and Wouter Saelens\textsuperscript{$\dagger{}$,1,2,4} \and Louise Deconinck\textsuperscript{1,2} \and Yvan Saeys\textsuperscript{1,2,*}}
\date{20 May 2021}

@@

\hypertarget{abstract}{%
	\section{Abstract}\label{abstract}}

We present dyngen, a multi-modal simulation engine for studying dynamic
cellular processes at single-cell resolution. dyngen is more flexible
than current single-cell simulation engines, and allows better method
development and benchmarking, thereby stimulating development and
testing of computational methods. We demonstrate its potential for
spearheading computational methods on three applications: aligning cell
developmental trajectories, cell-specific regulatory network inference
and estimation of RNA velocity.

\hypertarget{introduction}{%
	\section{Introduction}\label{introduction}}

@@

Another, more experimental use of simulation engines is their important
role in spearheading the development of computational methods, possibly
even before real data is available. In this way, simulation engines can
be used to assess the value of novel experimental protocols or
treatments. Simulation engines are also increasingly important when it
comes to finding alternatives to animal models, for example for drug
testing and precision medicine. In such scenarios, cellular simulations
can act as digital twins, offering unlimited experimentation \emph{in
	silico} \cite{bjornsson_digitaltwinspersonalize_2019}.

Simulating realistic data requires that the underlying biology is
recapitulated as best as possible, and in the case of transcriptomics
data this typically involves modelling the underlying gene regulatory
networks. Simulators of ``bulk'' microarray or RNA-sequencing profiles
simulate biological processes (e.g.~transcription, translation) by
translating a database of known regulatory interactions into a set of
ordinary differential equations (ODE)
\cite{roy_systemgeneratingtranscription_2008,hache_gengesystematicgeneration_2009,schaffter_genenetweaversilicobenchmark_2011,vandenbulcke_syntrengeneratorsynthetic_2006}.
These methods have been instrumental in performing benchmarking studies
\cite{prill_rigorousassessmentsystems_2010,marbach_revealingstrengthsweaknesses_2010,marbach_wisdomcrowdsrobust_2012}.
However, the advent of single-cell omics introduced several new types of
analyses (e.g.~trajectory inference, RNA velocity, cell-specific network
inference) which exploit the higher resolution of single-cell versus
bulk omics \cite{luecken_currentbestpractices_2019}. In addition,
the data characteristics of single-cell omics are vastly different from
bulk omics, typically having much lower library sizes and a higher
dropout rate, but also a high number of profiles
\cite{vallejos_normalizingsinglecellrna_2017}. The low library
sizes, in particular, are problematic as ODEs are ill-suited for
performing low-molecule simulations
\cite{gillespie_exactstochasticsimulation_1977}. This necessitates
the development of single-cell simulators.

@@

\begin{figure}[t!]
	\centering
	\includegraphics[width=\linewidth]{figure_1.pdf}
	\caption{
		\textbf{Showcase of dyngen functionality.}
		\textbf{A:} Changes in abundance levels are driven strictly by gene regulatory reactions.
		\textbf{B:} The input Gene Regulatory Network (GRN) is defined such that it models a dynamic process of interest.
		\textbf{C:} The reactions define how abundance levels of molecules change at any particular time point.
		\textbf{D:} Firing many reactions can significantly alter the cellular state over time.
		\textbf{E:} dyngen keeps track of the likelihood of a reaction firing during small intervals of time, called the propensity, as well as the actual number of firings.
		\textbf{F:} Similarly, dyngen can also keep track of the regulatory activity of every interaction.
		\textbf{G:} A benchmark of trajectory inference methods has already been performed using the cell state ground-truth \cite{saelens_comparisonsinglecelltrajectory_2019}.
		\textbf{H:} The cell state ground-truth enables evaluating trajectory alignment methods.
		\textbf{I:} The reaction propensity ground-truth enables evaluating RNA velocity methods.
		\textbf{J:} The cellwise regulatory network ground-truth enables evaluating cell-specific gene regulatory network inference methods.
	}
	\label{fig:overview}
\end{figure}

\hypertarget{results}{%
	\section{Results}\label{results}}

@@

\begin{figure}[t!]
	\centering
	\includegraphics[width=\linewidth]{figure_2.pdf}
	\caption{
		\textbf{dyngen provides ground-truth data for a variety of applications (left), which can be used to quantitatively evaluate methods (right).} Box plots denote the $Q_0$ to $Q_4$ quartile values.
		\textbf{A:} Trajectory alignment aligns two trajectories between samples. We evaluate dynamic time warping (DTW) and cellAlign when aligning two linear trajectories with different kinetic parameters based on the area differences between the worst possible alignment and the predicted alignment (Area Between Worst And Prediction, or ABWAP). 
		\textbf{B:} RNA velocity calculates for each cell the direction in which the expression of each gene is moving. We evaluated scVelo and velocyto by comparing these vectors with the known velocity vector (velocity correlation) and with the known direction of the cellular trajectory in a dimensionality reduction (velocity arrow cosine).
		\textbf{C:} Cell-specific network inference (CSNI) predicts the regulatory network of every individual cell. We evaluate each cell-specific regulatory network with typical metrics for network inference: the Area Under the Receiver Operating Characteristics-curve (AUROC) and Area Under the Precision Recall-curve (AUPR). We evaluate three CSNI methods by computing the mean AUROC and AUPR across all cells.
	}
	\label{fig:applications}
\end{figure}

@@

\hypertarget{discussion}{%
	\section{Discussion}\label{discussion}}

dyngen's single-cell simulations can be used to evaluate common
single-cell omics computational methods such as clustering, batch
correction, trajectory inference, and network inference. However, the
framework is flexible enough to be adaptable to a broad range of
applications, including methods that integrate clustering, network
inference, and trajectory inference. In this respect, dyngen may promote
the development of tools in the single-cell field similarly as other
simulators have done in the past
\cite{schaffter_genenetweaversilicobenchmark_2011,ewing_combiningtumorgenome_2015}.
Additionally, one could anticipate technological developments in
single-cell multi-omics. In this way, dyngen allows designing and
evaluating the performance and robustness of new types of computational
analyses before experimental data becomes available, comparing which
experimental protocol is the most cost-effective in producing
qualitative and robust results in downstream analysis. One major
assumption of dyngen is that cells are assumed to be well-mixed and
independent from each other. Subdividing a cell into multiple 2D or 3D
subvolumes or allowing cells to exchange molecules, respectively, could
pave the way to better study key cellular processes such as cell
division, intercellular communication, and migration
\cite{smith_spatialstochasticintracellular_2019}.

@@

\hypertarget{sec:dyngen-methods}{%
	\section{Methods}\label{sec:dyngen-methods}}

@@

\hypertarget{sec:dyngen-modules}{%
	\subsection{Defining the module network}\label{sec:dyngen-modules}}

@@

\hypertarget{sec:dyngen-grn}{%
	\subsection{Generating the gene regulatory
		network}\label{sec:dyngen-grn}}

@@

\hypertarget{sec:dyngen-reactions}{%
	\subsection{Convert gene regulatory network to a set of
		reactions}\label{sec:dyngen-reactions}}

@@

\hypertarget{propensity-of-transcription-when-n1}{%
	\subsubsection{\texorpdfstring{Propensity of transcription when
			\(N=1\)}{Propensity of transcription when N=1}}\label{propensity-of-transcription-when-n1}}

@@

\hypertarget{propensity-of-transcription-when-n2}{%
	\subsubsection{\texorpdfstring{Propensity of transcription when
			\(N=2\)}{Propensity of transcription when N=2}}\label{propensity-of-transcription-when-n2}}

@@

\hypertarget{propensity-of-transcription-for-an-arbitrary-n}{%
	\subsubsection{\texorpdfstring{Propensity of transcription for an
			arbitrary
			\(N\)}{Propensity of transcription for an arbitrary N}}\label{propensity-of-transcription-for-an-arbitrary-n}}

@@

\hypertarget{propensity-of-transcription-for-a-large-n}{%
	\subsubsection{\texorpdfstring{Propensity of transcription for a large
			\(N\)}{Propensity of transcription for a large N}}\label{propensity-of-transcription-for-a-large-n}}

@@

\hypertarget{independence-synergism-and-basal-expression}{%
	\subsubsection{Independence, synergism and basal
		expression}\label{independence-synergism-and-basal-expression}}

@@

\hypertarget{sec:dyngen-simcell}{%
	\subsection{Simulate single cells}\label{sec:dyngen-simcell}}

@@

\hypertarget{sec:dyngen-experiment}{%
	\subsection{Simulate experiment}\label{sec:dyngen-experiment}}

From the SSA simulation we obtain the abundance levels of all the
molecules at every state. We need to replicate technical effects
introduced by experimental protocols in order to obtain data that is
similar to real data. For this, the cells are sampled from the
simulations and molecules are sampled for each of the cells. Gene
capture rates and library sizes are empirically derived from real
datasets to match real technical variation.

\hypertarget{sample-cells}{%
	\subsubsection{Sample cells}\label{sample-cells}}

@@

\hypertarget{sample-molecules}{%
	\subsubsection{Sample molecules}\label{sample-molecules}}

@@

\hypertarget{comparison-between-a-dyngen-and-a-reference-dataset}{%
	\subsubsection{Comparison between a dyngen and a reference
		dataset}\label{comparison-between-a-dyngen-and-a-reference-dataset}}

Comparison between a dyngen dataset and the reference dataset it used in
terms of characteristic single-cell omics features showed that dyngen
produces datasets with highly similar data characteristics
(Supplementary Note 1). Supplementary Note 1 was generated using
countsimQC \cite{soneson_unifiedqualityverification_2018}. .

\hypertarget{sec:dyngen-batcheffect}{%
	\subsection{Simulating batch effects}\label{sec:dyngen-batcheffect}}

@@

\hypertarget{sec:dyngen-groundtruth}{%
	\subsection{Determining the ground-truth
		trajectory}\label{sec:dyngen-groundtruth}}

@@

\hypertarget{sec:dyngen-extractgrn}{%
	\subsection{Determining the cell-specific ground-truth regulatory
		network}\label{sec:dyngen-extractgrn}}

Calculating the regulatory effect of a regulator \(R\) on a target \(T\)
(Supplementary Fig.~4F) requires determining the contribution of \(R\)
in the propensity function of the transcription of \(T\) with respect to
other regulators. This information is useful, amongst others, for
benchmarking cell-specific network inference methods.

@@

\hypertarget{sec:dyngen-nicompare}{%
	\subsection{Comparison of cell-specific network inference
		methods}\label{sec:dyngen-nicompare}}

42 datasets were generated using the 14 different predefined backbones
and three different seeds. For every cell in the dataset, the
transcriptomics profile and the corresponding cell-specific ground-truth
regulatory network was determined.

@@

\hypertarget{sec:dyngen-velcompare}{%
	\subsection{Comparison of RNA velocity
		methods}\label{sec:dyngen-velcompare}}

@@

\hypertarget{sec:dyngen-tacompare}{%
	\subsection{Comparison of trajectory
		alignment}\label{sec:dyngen-tacompare}}

@@

\hypertarget{comparison-of-scalability-and-runtime}{%
	\subsection{Comparison of scalability and
		runtime}\label{comparison-of-scalability-and-runtime}}

Simulating a bifurcating cycle dataset with 10'000 genes and 10'000
required in total 1147 seconds (Supplementary Note 2). Fixing the number
of genes and varying the number of cells showed that the execution time
of dyngen scales linearly w.r.t. the number of cells (Supplementary Note
2). Fixing the number of cells and varying the number of genes also
showed that the execution time of dyngen scales linearly w.r.t. the
number of genes (Supplementary Note 2). These timings were measured
using 30 (out of 32) threads using a AMD Ryzen 9 5950X clocked at
3.4GHz.

\hypertarget{data-availability}{%
	\section{Data Availability}\label{data-availability}}

@@

\hypertarget{code-availability}{%
	\section{Code Availability}\label{code-availability}}

@@

\hypertarget{acknowledgements}{%
	\section{Acknowledgements}\label{acknowledgements}}

This project has been made possible in part by grant number 2020-218899
from the Chan Zuckerberg Initiative DAF, an advised fund of Silicon
Valley Community Foundation. This research received funding from the
Flemish Government under the ``Onderzoeksprogramma Artificiële
Intelligentie (AI) Vlaanderen'' program.

\hypertarget{author-contributions}{%
	\section{Author contributions}\label{author-contributions}}

@@

\hypertarget{competing-interests}{%
	\section{Competing interests}\label{competing-interests}}

The authors declare no competing interests.

%\bibliographystyle{unsrtnat}
%\bibliography{../library}

\begin{thebibliography}{59}
	\providecommand{\natexlab}[1]{#1}
	\providecommand{\url}[1]{\texttt{#1}}
	\expandafter\ifx\csname urlstyle\endcsname\relax
	\providecommand{\doi}[1]{doi: #1}\else
	\providecommand{\doi}{doi: \begingroup \urlstyle{rm}\Url}\fi

	\bibitem[Zappia et~al.(2017)Zappia, Phipson, and
	Oshlack]{zappia_splattersimulationsinglecell_2017}
	Luke Zappia, Belinda Phipson, and Alicia Oshlack.
	\newblock Splatter: {{Simulation}} of {{Single}}-{{Cell RNA Sequencing Data}}.
	\newblock \emph{Genome Biology}, 18:\penalty0 174, September 2017.
	\newblock ISSN 1474-760X.
	\newblock \doi{10.1186/s13059-017-1305-0}.

	\bibitem[Bj{\"o}rnsson et~al.(2019)Bj{\"o}rnsson, Borrebaeck, Elander,
	Gasslander, Gawel, Gustafsson, J{\"o}rnsten, Lee, Li, Lilja,
	{Mart{\'i}nez-Enguita}, Matussek, Sandstr{\"o}m, Sch{\"a}fer, Stenmarker,
	Sun, Sysoev, Zhang, Benson, and {on behalf of the Swedish Digital Twin
		Consortium}]{bjornsson_digitaltwinspersonalize_2019}
	Bergthor Bj{\"o}rnsson, Carl Borrebaeck, Nils Elander, Thomas Gasslander,
	Danuta~R. Gawel, Mika Gustafsson, Rebecka J{\"o}rnsten, Eun~Jung Lee, Xinxiu
	Li, Sandra Lilja, David {Mart{\'i}nez-Enguita}, Andreas Matussek, Per
	Sandstr{\"o}m, Samuel Sch{\"a}fer, Margaretha Stenmarker, X.~F. Sun, Oleg
	Sysoev, Huan Zhang, Mikael Benson, and {on behalf of the Swedish Digital Twin
		Consortium}.
	\newblock Digital twins to personalize medicine.
	\newblock \emph{Genome Medicine}, 12\penalty0 (1):\penalty0 4, December 2019.
	\newblock ISSN 1756-994X.
	\newblock \doi{10.1186/s13073-019-0701-3}.

	\bibitem[Roy et~al.(2008)Roy, {Werner-Washburne}, and
	Lane]{roy_systemgeneratingtranscription_2008}
	Sushmita Roy, Margaret {Werner-Washburne}, and Terran Lane.
	\newblock A system for generating transcription regulatory networks with
	combinatorial control of transcription.
	\newblock \emph{Bioinformatics}, 24\penalty0 (10):\penalty0 1318--1320, May
	2008.
	\newblock ISSN 1367-4803.
	\newblock \doi{10.1093/bioinformatics/btn126}.

	\bibitem[Hache et~al.(2009)Hache, Wierling, Lehrach, and
	Herwig]{hache_gengesystematicgeneration_2009}
	Hendrik Hache, Christoph Wierling, Hans Lehrach, and Ralf Herwig.
	\newblock {{GeNGe}}: Systematic generation of gene regulatory networks.
	\newblock \emph{Bioinformatics}, 25\penalty0 (9):\penalty0 1205--1207, May
	2009.
	\newblock ISSN 1367-4803.
	\newblock \doi{10.1093/bioinformatics/btp115}.

	\bibitem[Schaffter et~al.(2011)Schaffter, Marbach, and
	Floreano]{schaffter_genenetweaversilicobenchmark_2011}
	Thomas Schaffter, Daniel Marbach, and Dario Floreano.
	\newblock {{GeneNetWeaver}}: In silico benchmark generation and performance
	profiling of network inference methods.
	\newblock \emph{Bioinformatics}, 27\penalty0 (16):\penalty0 2263--2270, August
	2011.
	\newblock ISSN 1367-4811.
	\newblock \doi{10.1093/bioinformatics/btr373}.

	\bibitem[{Van den Bulcke} et~al.(2006){Van den Bulcke}, Van~Leemput, Naudts,
	{van Remortel}, Ma, Verschoren, De~Moor, and
	Marchal]{vandenbulcke_syntrengeneratorsynthetic_2006}
	Tim {Van den Bulcke}, Koenraad Van~Leemput, Bart Naudts, Piet {van Remortel},
	Hongwu Ma, Alain Verschoren, Bart De~Moor, and Kathleen Marchal.
	\newblock {{SynTReN}}: A generator of synthetic gene expression data for design
	and analysis of structure learning algorithms.
	\newblock \emph{BMC Bioinformatics}, 7\penalty0 (1):\penalty0 43, January 2006.
	\newblock ISSN 1471-2105.
	\newblock \doi{10.1186/1471-2105-7-43}.

	\bibitem[Prill et~al.(2010)Prill, Daniel, Julio, K., G., Xiaowei, D., Gregoire,
	Gustavo, Marbach, {Saez-Rodriguez}, Sorger, Alexopoulos, Xue, Clarke,
	{Altan-Bonnet}, Stolovitzky, Daniel, Julio, K., G., Xiaowei, D., Gregoire,
	and Gustavo]{prill_rigorousassessmentsystems_2010}
	Robert~J Prill, Marbach Daniel, Saez-Rodriguez Julio, Sorger~Peter K.,
	Alexopoulos~Leonidas G., Xue Xiaowei, Clarke~Neil D., Altan-Bonnet Gregoire,
	Stolovitzky Gustavo, Daniel Marbach, Julio {Saez-Rodriguez}, Peter~K Sorger,
	Leonidas~G Alexopoulos, Xiaowei Xue, Neil~D Clarke, Gregoire {Altan-Bonnet},
	Gustavo Stolovitzky, Marbach Daniel, Saez-Rodriguez Julio, Sorger~Peter K.,
	Alexopoulos~Leonidas G., Xue Xiaowei, Clarke~Neil D., Altan-Bonnet Gregoire,
	and Stolovitzky Gustavo.
	\newblock Towards a {{Rigorous Assessment}} of {{Systems Biology Models}}:
	{{The DREAM3 Challenges}}.
	\newblock \emph{PLoS ONE}, 5\penalty0 (2):\penalty0 e9202, January 2010.
	\newblock ISSN 1932-6203.
	\newblock \doi{10.1371/journal.pone.0009202}.

	\bibitem[Marbach et~al.(2010)Marbach, Prill, Schaffter, Mattiussi, Floreano,
	and Stolovitzky]{marbach_revealingstrengthsweaknesses_2010}
	Daniel Marbach, Robert~J Prill, Thomas Schaffter, Claudio Mattiussi, Dario
	Floreano, and Gustavo Stolovitzky.
	\newblock Revealing strengths and weaknesses of methods for gene network
	inference.
	\newblock \emph{Proceedings of the National Academy of Sciences}, 107\penalty0
	(14):\penalty0 6286--6291, April 2010.
	\newblock ISSN 1091-6490.
	\newblock \doi{10.1073/pnas.0913357107}.
